---
title: "Excitatory and Inhibitory Dynamics of Model Neurons"
subtitle: "A Study By Wilson, H. R. and Cowan, J. D"
author: "Hannah Luebbering"
date: December 12, 2020
output:
  github_document: default
---


```{r include=FALSE}
library(kableExtra)
library(deSolve)
library(graphics)
library(mathjaxr)
library(pagedown)
library(captioner)
library(latex2exp)
knitr::opts_chunk$set(echo = FALSE, out.width = 400, fig.align = "center", fig.pos = 'default', fig.show = "hold", warning = FALSE, message = FALSE)
library(phaseR)
library(knitr)
library(nat)
library(rmarkdown)
```




#### Limit Cycles


Non-linear systems can exhibit a type of behavior known as a limit cycle. If there is only one steady-state solution, and if the steady-state solution is unstable, a limit cycle will occur. In the following, we define the parameters for satisfying conditions 18 and 20 as $c_1=16$, $c_2 = 12$, $c_3=15$, $c_4=3$, $a_e = 1.3$, $\theta_e=4$, $a_i=2$, $\theta_i = 3.7$, $r_e=1$ and $r_i=1$. We can determine a steady-state solution by the intersection of the nullclines as follows.


```{r}
fig_nums <- captioner(prefix = "Figure")
```


```{r}
se <- function(x){
  ae=1.3
  theta_e=4
  1/(1+exp(-ae*(x-theta_e)))
}

si <- function(x){
  ai=2
  theta_i= 3.7
  1/(1+exp(-ai*(x-theta_i)))
}

```



```{r}
WilsonCowan2 <- function(t, y, parameters) {
  # couplings
  c1 = 16
  c2 = 12
  c3 = 15
  c4 = 3
  
  # Refractory periods
  rE = 1
  rI = 1
  
  # external inputs
  P = 1.25
  Q = 0
  
  ki=0.825
  ke=0.88
  I <- y[1]
  E <- y[2]
  dy <- c(
     -I + (ki - rI * I) * si(c3 * E - c4 * I + Q),
     -E + (ke - rE * E) * se(c1 * E - c2 * I + P))
  list(dy)
}
```



```{r}
body_cap8 <- fig_nums(name = "phase8", caption = "Phase Plane Analysis. Determine the steady-state solution by the nullclines' intersection. Parameters: $c_1=16$, $c_2 = 12$, $c_3=15$, $c_4=3$, $a_e = 1.3$, $\\theta_e=4$, $a_i=2$, $\\theta_i = 3.7$, $r_e=1$, $r_i=1$.")
```



```{r, fig.cap=body_cap8, fig.show='hold', message=FALSE}
example4_flowField  <- flowField(WilsonCowan2,
                                 xlim = c(-0.01, .425),
                                 ylim = c(0, .5),
                                 add  = FALSE,
                                 ylab = TeX("$E$"),
                                 xlab= TeX("$I$"),
                                 frac=1)
grid()
example4_nullclines <- nullclines(WilsonCowan2,
                                  xlim = c(-0.01, .425),
                                  ylim = c(0, .5),
                                  lty = 2, lwd = 2,
                                  col=c("lightseagreen","aquamarine4"))

```





In the phase plane, the limit cycle is an isolated closed orbit, where "closed" means the periodicity of movement, and "isolated" means the limit of motion, where nearby trajectories converge or deviate. We can alter our initial values of $E_0$ and $I_0$ to obtain different paths in the phase space.


```{r}
body_cap9 <- fig_nums(name = "phase9", caption = "Phase Plane Analysis showing limit cycle trajectory in response to constant simulation $P=1.25$. Dashed lines are nullclines. Parameters: $c_1=16$, $c_2 = 12$, $c_3=15$, $c_4=3$, $a_e = 1.3$, $\\theta_e=4$, $a_i=2$, $\\theta_i = 3.7$, $r_e=1$, $r_i=1$.")
```



```{r Final, fig.cap=body_cap9, fig.show='hold', message=FALSE}
example4_flowField  <- flowField(WilsonCowan2,
                                 xlim = c(-0.01, .425),
                                 ylim = c(0, .5),
                                 add  = FALSE,
                                 ylab = TeX("$E$"),
                                 xlab= TeX("$I$"),
                                 frac=1)
grid()
example4_nullclines <- nullclines(WilsonCowan2,
                                  xlim = c(-0.01, .425),
                                  ylim = c(0, .5),
                                  lty = 2, lwd = 2,
                                  col=c("lightseagreen","aquamarine4"))

y0 <- matrix(c(0.1,0.3,
               0,0,
               0.1,0.2), 3, 2, byrow = TRUE)
example4_trajectory <- trajectory(WilsonCowan2,
                                  y0   = y0,
                                  pch=16,
                                  tlim = c(0, 100),
                                  col="black",
                                  add=T, ylab=TeX("$E, I$"), xlab=TeX("$t$"))
grid()
```


The phase plane analysis illustrates a bounded steady-state solution that is classified as unstable; this is a typical feature of a limit cycle. The solution's oscillating behavior, shown in Figure 10, follows typical limit cycle behavior:

<span class="lp6">i.</span> Trajectories near the equilibrium point are pushed further away from the equilibrium.

<span class="lp6">ii.</span> Trajectories far from the equilibrium point move closer toward the equilibrium.



We established the resting state $E=0, I=0$ to be stable in the absence of an outside force. Therefore, the neural population only exhibits limit cycle activity in response to a constant external input (P or Q). All in all, the premise of Theorem 3 is that there is only one steady-state, and it must be near the inflection point for a limit cycle to exist. Therefore, if we study the limit behavior as a function of $P$, where $Q=0$, then it follows that: 

- There is a threshold value of P, and below this threshold, the limit cycle activity cannot occur.
- There is a higher value of P, and above this bound, the system's limit cycle activity will cease.
- Within the range defined above, both the limit cycle frequency and the average value of $E(t)$ increases monotonically with respect to $P$.



```{r}
body_cap10 <- fig_nums(name = "phase10", caption = "$I(t)$ and $E(t)$ for limit cycle shown in Fig. 9. The limit cycle depends on the value of $P$, i.e. $Q$ being set equal to zero.")
```


\ 


```{r echo=FALSE, fig.cap=body_cap10, fig.show='hold', out.width=350 , message=FALSE, warning=FALSE}

example4_solution <- numericalSolution(WilsonCowan2,
                        y0 = c(0.0, 0.1),
                        type = "two",
                        col=c("cornflowerblue", "aquamarine4"),
                        add.legend = T,
                        xlab = TeX("$t$"),
                        ylab = c(TeX("$I$"), TeX("$E$")),
                        add.grid = F,
                        tlim = c(0,20),
                        lwd=2,
                        ylim=c(0,0.3),
                        xlim=c(0,18))
```





