---
output: 
  html_document: 
    toc: yes
    toc_float: yes
    css: static/phase-r.css
bibliography: static/citation/phase-r-citation.bib
nocite: '@*'
---

## Phase Plane Analysis in R



`phaseR`: An R package for phase plane analysis of one and two-dimensional autonomous ODE systems [@grayling2014].



------------------------------



## Example 1. one-dimesnional ODE


<div class="a">
Consider the one-dimensional autonomous ODE:
$$\frac{dy}{dt}=y(1-y)(2-y)$$

</div>



### Flow Field

Consider the one-dimensional autonomous ODE:
$$\frac{dy}{dt}=y(1-y)(2-y)$$

$$\longrightarrow \hat y= \frac{dy}{dt}$$


The following plots the flow field and various trajectories, adding **horizontal lines at equilibrium points**:


```{r echo=FALSE, message=FALSE, warning=FALSE, out.width=350}
library(phaseR)

example2_flowField  <- flowField(
  example2, 
  xlim = c(0, 4), 
  ylim = c(-1, 3), 
  system ="one.dim", 
  add = FALSE)

grid()

example2_nullclines <- nullclines(
  example2, xlim = c(0, 4), ylim = c(-1, 3), system = "one.dim")

example2_trajectory <- trajectory(
  example2, 
  y0 = c(-0.5, 0.5, 1.5, 2.5), 
  tlim = c(0, 4), 
  system = "one.dim")

```


### Find Fixed Points


The horizontal lines on the graph indicate that **3 equilibrium points have been identified at $y^*=0,1,2$**

If we set $\hat y=0,$ we can analytically solve for the 3 equilibrium points:

$$ y^* (1-y^* )(2-y^* )=0 \\
\Longrightarrow y^*=0,1,2 $$


### Determine Stability of Fixed Points

##### Method 1. `Phase Portrait`

<div class="b1">

Plotting the **phase portrait**, we find that $y^*=0$ and $y^*=2$ are unstable; and $y^*=1$ is stable 

</div>

\ 

```{r, echo = F, message=FALSE, warning=FALSE, fig.dim = c(7, 5)}
example2_phasePortrait <- phasePortrait(
  example2, ylim = c(-0.5, 2.5))
```



------------------------------


##### Method 2. `Taylor Series Approach`

<div class="b1">

Using the Taylor Series approach we have:

$$ \frac{d}{dy}\left.\left(\frac{dy}{dt}\right)\right|_{y=y^*} = 3y^{*^2} - 6y^* + 2 = \begin{cases} 2 &: \ y^*=0,\\ -1 &: \ y^*=1,\\ 2 &: \ y^*=2.  \end{cases} $$

</div>

\ 

We draw the same conclusion as from the phase portrait. We can confirm the Taylor analysis using `stability()` to check the **stability of each equilibrium point**:


```{r, echo = F}
example2_stability_1 <- stability(
  example2, ystar = 0, system = "one.dim")

example2_stability_2 <- stability(
  example2, ystar = 1, system = "one.dim")

example2_stability_3 <- stability(
  example2, ystar  = 2, system = "one.dim")

```



------------------------------


##### Perturbation Analysis 

* assuming higher order terms can be neglected

Note $f(y^*)=0$, the autonomous ODE becomes:

$$ \frac{d}{dt}(y^*+\delta)=\delta\frac{\partial f}{\partial y}(y^*) \Rightarrow \frac{d\delta}{dt}=\delta\frac{\partial f}{\partial y}(y^*)=k\delta. $$

The above ODE for $\delta$ can be solved to give $\delta(t)=\delta(0)e^{kt}$. So, stability can be found depending on whether $\delta(t)$ grows or decays as $t$ increases:

$$ k = \frac{\partial f}{\partial y}(y^*)\begin{cases} > 0 &: \ \text{Stable},\\ < 0 &: \ \text{Unstable}.  \end{cases} $$

Here, $k$ can be called the discriminant


-------------------------------

Returning to the given example...

$$ \frac{\partial f}{\partial y}(y^*) = -2y^*= \begin{cases} -4 &: \ y^*=2, \\ +4 &: \ y^* = -2. \end{cases}$$

Hence we reach conclusion as above; 

* $y^*=2$ is stable, and $y^*=-2$ is unstable 

The above shows that we can state if $y(0)>2$ or $0<y(0)<2$ then the solution will eventually approach $y=2$. 

However, if $y(0)<0$, $y\rightarrow-\infty$ as $t\rightarrow\infty$. 


----------------------

##### Taylor Series Conditions

If $\mathcal{k}=0$,

$$\Rightarrow\frac{\partial f}{\partial y}(y^*)=0, $$

then to the order of the Taylor Series **no conclusion can be drawn about stability**. 


----------------------


### Summary

The above demonstrates the components necessary to perform qualitative analysis on a one-dimensional autonomous ODE. 

`General Process:`

<div class="a">

<ol>
<li>plot the flow field</li>
<li>plot several trajectories one the flow field</li>
<li>identify the equilibrium points</li>
<li>choose a method to determine stability of equilibrium points</li>
</ol>

</div>




---------------------------------------



## Example 2. Logistic Growth Model


<div class="b1">

The logistic growth model is used in biology to model the growth of a population under density dependence. 

$$ \frac{dy}{dt} = \beta y\left(1-\frac{y}{K}\right). $$

</div>

\ 

Letting $\beta=1$ and $K=2$, the following plots the flow field and several trajectories, along with added horizontal lines to indicate equilibrium points.


```{r echo=FALSE, out.width=300, message=FALSE}
library(phaseR)
logistic_flowField  <- flowField(
  logistic,
  xlim= c(0, 5),
  ylim= c(-1, 3),
  parameters = c(1, 2),
  system= "one.dim",
  add= FALSE)

grid()

logistic_nullclines <- nullclines(
  logistic,xlim= c(0, 5),ylim= c(-1, 3),
  parameters = c(1, 2),
  system= "one.dim")

logistic_trajectory <- trajectory(
  logistic,y0= c(-0.5, 0.5, 1.5, 2.5),
  tlim= c(0, 5),
  parameters = c(1, 2),
  system= "one.dim")

```


<div class="a">

From the horizontal lines graphed above, we can identify 2 equilibrium points at $y=0 \,\,\text{and } y=K$

</div>

\ 

<div class="b1">

We can confirm these equilibrium points by analytically solving for $\dot y =0$


$$ \beta y^* \left(1-\frac{y^*}{K}\right)=0 \\ \Longrightarrow y^*=0,\;K $$

</div>



---------------------------

Next, we plot the **phase portrait**:


```{r echo=F, out.width=300, message=FALSE}
logistic_phasePortrait <- phasePortrait(
  logistic,
  ylim= c(-0.5, 2.5),
  parameters = c(1, 2))

```

<div class="a">

From the phase portrait, we can observe that **$y^*=0$ is unstable and $y^*=K$ stable**, for the case $\beta=1$ and $K=2$.

</div>



-----------------------------

If we use `Taylor Series method` we can draw the same conclusion:

<div class="b1">

$$ \frac{d}{dy}\left.\left(\frac{dy}{dt}\right)\right|_{y=y^*} = \beta - \frac{2\beta y^*}{K} = \begin{cases} \beta &: \ y^*=0,\\ -\beta &: \ y^*=K.  \end{cases} $$



So for $\beta=1$ and $K=2$, we have a stable point at $y=2$. So, we can see that the point $y=K$ will in general be stable provided $\beta>0$.

</div>

\ 


The following code verifies the above results for the stabilities of the equilibrium points:


```{r, echo = T}
## equilibrium point y*=0
logistic_stability_1 <- stability(
  logistic,
  ystar= 0,
  parameters = c(1, 2),
  system= "one.dim")

## equilibrium point y*=K
logistic_stability_2 <- stability(
  logistic,
  ystar= 2,
  parameters = c(1, 2),
  system= "one.dim")

```




-------------------------

### References




